Choosing Color Palettes for Statistical Graphics

### 2007-11-06 pkg/paper/Palettes/CSDA.Rnw
hack and slash on the intro to CSDA
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@23 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

\paragraph{The colors should not be unappealing.}
It is not necessary for the 
colors in a statistical plot to reflect the latest in fashion trends, 
but basic 
principles such as avoiding fully saturated colors should be
adhered to.  The requirement is not that the user should have
a degree in graphic design, but
that the software should allow the user to select colors and provide
control over basic properties such as how bright the colors are.
In order to be able to control basic perceptual features of colors,
it is necessary to work in a \emph{colorspace} that describes 
colors in terms of their perceptual properties.
This article will describe an appropriate colorspace for
selecting colors.

\paragraph{The colors in a statistical graphic should cooperate with each other.}
The typical purpose 
of color in a statistical graphic is to distinguish between different 
areas or symbols in the plot---to distinguish between different
groups or between different levels of a variable.  This means
that there will typically be several colors, or a \emph{palette}
of colors, used within a plot and 
that those colours should be related to each other.
Part of the solution to this problem is to work in an
appropriate colorspace, as in the previous point, so that 
motion within the colorspace and distances between colors 
have an intuitive and rational meaning.  However, 
there still remains the issue of providing basic navigation 
tools so that users can conveniently generate color
palettes.  The main part of this article will describe
algorithms for generating color palettes for common 
applications in statistical graphics.

\paragraph{The colors should work everywhere.}
In the ideal situation, colors 
should be selected so that they continue to work in any 
context.  For example, different areas of a plot should still be 
distiguishable when printed in black-and-white, or when the person
viewing the graphic is colour-blind.
This article will also briefly address these issues.
%FIXME% Also mention that the colors should work on different devices.
%FIXME% Explain that we discuss how and under which circumstances this
%FIXME% is achieved.

@@

For implementing color palettes in computer programs, typically color spaces are
employed that provide a mapping to Red-Green-Blue (RGB) colors. RGB is an additive
color model which is used for generating colors on computer screens
\citep[by mixing different amounts of the primary colors red, green and blue,
see][for more details]{color:Poynton:2000}. While being eminently useful for generating
colors on computer screens, it is hard for humans to select the coordinates in RGB
space corresponding to a certain color. Therefore, color picker tools in computer
programs are based on different color models---starting from the pioneering introduction
of Hue-Saturation-Value (HSV) colors \citep{color:Smith:1978}---which try to capture the
dimensions of human visual perception. HSV space is a simple transformation of RGB
space which is also implemented in many statistical software packages
\citep[see][for more details on HSV colors and links to further information about the other color
spaces discussed in the following]{color:Wiki+HSV:2007}. Although being easy to use,
HSV space exhibits a number of disadvantages: It is relatively difficult to select sets of
HSV coordinates that yield colors that are ``in harmony'' \citep[see][]{color:Munsell:1905}.
The reason for this is that colors with different hues also differ in saturation and brightness.
For statistical graphics, this is important because it can introduce size distortions in
the perception of shaded areas and color-caused optical illusions \citep{color:Cleveland+McGill:1983}.
Furthermore, the use of HSV colors encourages the use of flashy and highly saturated
colors which are good for drawing attention to a plot but hard to look at for a longer time.

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{color:Harrower+Brewer:2003},
an online tool for selecting color schemes for maps. It provides a rich
collection of prefabricated palettes (with a fixed maximal number
of colors) and guides its users how to choose a suitable palette for coding
various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{color:Brewer:1999} and \cite{color:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables
(or ordinal variables if some numerical coding for its levels is used).
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
a general principle for selecting colors by traversing paths along perceptual axes in
a suitable color space. Consequently, the user can decide which path exactly should be
taken and how many colors should be selected. The color space employed by us is the
Hue-Chroma-Luminance (HCL) color space \citep[see][]{color:Ihaka:2003} and we show how
the general principles suggested can be turned into formulas describing paths of
coordinates in this space.
%% Given a mapping from HCL to RGB coordinates, our formulas are
%% extremely easy to implement in software and we provide such an implementation
%% in the \proglang{R} language \citep{color:R:2007} using the powerful \proglang{R} graphics
%% system \citep[see][]{color:Murrell:2006} and the HCL
%% color implementation from package \pkg{colorspace} \citep{color:Ihaka:2006}. This gives the user
%% both a conceptual and computational tool box for experimenting with color palettes for
%% a particular display.

The remainder of the paper is organized as follows: Section~\ref{sec:motivation} provides
several motivating examples, showing how typical HSV-based graphics can be enhanced
by using HCL-based palettes. Section~\ref{sec:spaces} gives a brief introduction
to the underlying HSV and HCL color spaces before Section~\ref{sec:palettes} 
suggests strategies for deriving HCL-based qualitative, sequential, and diverging palettes.
Section~\ref{sec:software} offers some general remarks on the implementation in statistical
software packages as well as some details on the implementation in \proglang{R}.
Section~\ref{sec:discussion} concludes the paper with a discussion.

@@

Figure~\ref{fig:pal-q-construct} depicts how the HCL-based color wheel is constructed.
It shows the hue/chroma plane of HCL space given a luminance of $L = 70$. Not all
combinations of hue and chroma are admissible, however for a chroma of $C = 50$ a full
color wheel can be obtained.
For choosing the hues in a certain palette, various strategies are conceivable.
A simple and intuitive one is to use colors as metaphors for categories
(e.g., for political parties as in Figures~\ref{fig:seats} and \ref{fig:votes}),
another approach would be to use segments from
the color wheel corresponding to nearby or distant colors. The latter is shown
in Figure~\ref{fig:pal-q-examples} which depicts
examples for generating qualitative sets of colors $(H, 50, 70)$.
In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.

>>>>

\usepackage{amsfonts}

@@

\paragraph{The colors should not be unappealing.}
It is not necessary for the 
colors in a statistical plot to reflect the latest in fashion trends, 
but basic 
principles such as avoiding fully saturated colors should be
adhered to.  The requirement is not that the user should have
a degree in graphic design, but
that the software should allow the user to select colors and provide
control over basic properties such as how bright the colors are.
In order to be able to control basic perceptual features of colors,
it is necessary to work in a \emph{colorspace} that describes 
colors in terms of their perceptual properties.

Typically, software allows the user to specify colors as 
Red-Green-Blue (RGB) triplets.  
RGB is an additive
color model which is used for generating colors on computer screens
\citep[by mixing different amounts of the primary colors red, green and blue,
see][for more details]{color:Poynton:2000}.
Unfortunately, it is impossible to control
the perceptual properties of a color in this colorspace because there
is no single dimension that corresponds to, for example, the brightness
of the colour.  In
this article we will describe a more appropriate colorspace for
selecting colors.

\paragraph{The colors in a statistical graphic should cooperate with each other.}
The typical purpose 
of color in a statistical graphic is to distinguish between different 
areas or symbols in the plot---to distinguish between different
groups or between different levels of a variable.  This means
that there will typically be several colors, or a \emph{palette}
of colors, used within a plot and 
that those colors should be related to each other.

Part of the solution to this problem is to work in an
appropriate colorspace, as in the previous point, so that 
motion within the colorspace and distances between colors 
have an intuitive and rational meaning.  
For example, another common software 
approach \citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004} 
is to allow
colors to be specified as Hue-Saturation-Value (HSV) 
triplets \citep{color:Smith:1978},
where each dimension controls a separate perceptual 
property of the color.  
HSV space is a simple transformation of RGB
space 
\citep[see][for more details on HSV colors and links to further information about the other color
spaces discussed in the following]{color:Wiki+HSV:2007} and,
unfortunately, the dimensions in 
HSV space are not truly independent.  
For example,
a change in hue results in a color that may also be
perceptually brighter and more saturated.  
This means that it is relatively difficult to select sets of
HSV coordinates that yield colors that are ``in harmony'' \citep[see][]{color:Munsell:1905}.
For statistical graphics, this is important because it can introduce size distortions in
the perception of shaded areas and color-caused optical illusions \citep{color:Cleveland+McGill:1983}.
Furthermore, the use of HSV colors encourages the use of flashy and highly saturated
colors which are good for drawing attention to a plot but hard to look at for a longer time.

A different approach to obtaining a color palette is provided by the 
\pkg{ColorBrewer.org} \citep{color:Harrower+Brewer:2003}
online tool for selecting color schemes for maps. It provides a rich
collection of prefabricated palettes (with a fixed maximal number
of colors) and guides its users how to choose a suitable palette for coding
various types of information.  The drawback to this tool is that
it only provides a fixed set of colors for each palette and there
is no way to extend the existing palettes.

Following \cite{color:Brewer:1999} and \cite{color:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables
(or ordinal variables if some numerical coding for its levels is used).
Unlike \pkg{ColorBrewer.org}, we not only provide fixed sets of colors but suggest
a general principle for selecting colors by traversing paths along perceptual axes in
a suitable color space. Consequently, the user can decide which path exactly should be
taken and how many colors should be selected. The color space we will work in
 is the
Hue-Chroma-Luminance (HCL) color space \citep[see][]{color:Ihaka:2003} and we show how
the general principles suggested can be turned into formulas describing paths of
coordinates in this space.

\paragraph{The colors should work everywhere.}
The final issue to deal with is that, in an ideal situation, colors 
should be selected so that they continue to work in any 
context.  For example, different areas of a plot should still be 
distiguishable when the person
viewing the graphic is color-blind
or when the graphic is printed in black-and-white instead of being
viewed on screen.
These goals cannot always be attained, but we will discuss 
situations where attention should be paid to these issues and
suggest strategies for resolving any problems.
%FIXME% Also mention that the colors should work on different devices.

@@

%% Given a mapping from HCL to RGB coordinates, our formulas are
%% extremely easy to implement in software and we provide such an implementation
%% in the \proglang{R} language \citep{color:R:2007} using the powerful \proglang{R} graphics
%% system \citep[see][]{color:Murrell:2006} and the HCL
%% color implementation from package \pkg{colorspace} \citep{color:Ihaka:2006}. This gives the user
%% both a conceptual and computational tool box for experimenting with color palettes for
%% a particular display.

\paragraph{}
The remainder of the paper is organized as follows: Section~\ref{sec:motivation} provides
several motivating examples, showing how typical HSV-based graphics can be enhanced
by using HCL-based palettes. Section~\ref{sec:spaces} gives a brief introduction
to the underlying HSV and HCL color spaces before Section~\ref{sec:palettes} 
suggests strategies for deriving HCL-based qualitative, sequential, and diverging palettes.
Section~\ref{sec:software} offers some general remarks on the implementation in statistical
software packages as well as some details on the implementation in \proglang{R}.
Section~\ref{sec:discussion} concludes the paper with a discussion.

@@

Figure~\ref{fig:pal-q-construct} depicts how the HCL-based color wheel is constructed.
It shows the hue/chroma plane of HCL space given a luminance of $L = 70$. Not all
combinations of hue and chroma are admissible, however for a chroma of $C = 50$ a full
color wheel can be obtained.
For choosing the hues in a certain palette, various strategies are conceivable.
%% FIXME:  the figures referred to below no longer exist!!
A simple and intuitive one is to use colors as metaphors for categories
(e.g., for political parties as in Figures~\ref{fig:seats} and \ref{fig:votes}),
another approach would be to use segments from
the color wheel corresponding to nearby or distant colors. The latter is shown
in Figure~\ref{fig:pal-q-examples} which depicts
examples for generating qualitative sets of colors $(H, 50, 70)$.
In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.
